---
output: github_document
---


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)

pkgload::load_all(".", export_all = FALSE, helpers = FALSE, quiet = TRUE)
ggplot2::theme_set(theme_funcml())
set.seed(42)
```



# funcml

`funcml` is a functional machine learning framework for tabular data in R.

It provides one explicit interface for the core modeling workflow:

* fit models with `fit()`
* generate predictions with `predict()`
* validate performance with `evaluate()`
* tune hyperparameters with `tune()`
* compare learners with `compare_learners()`
* interpret fitted models with `interpret()`
* estimate causal effects with `estimate()`

The package is intentionally compact and opinionated: preprocessing happens before modeling, inputs stay explicit, and the API stays small instead of expanding into a large orchestration layer.

## Installation

```r
install.packages("remotes")
remotes::install_github("ielbadisy/funcml")
```

## Core API

The design of `funcml` centers on a small set of functions:

```r
fit()
predict()
evaluate()
tune()
compare_learners()
interpret()
estimate()
```

## Explore the registry

`funcml` exposes a session-aware registry of learners, metrics, and interpretation methods.

```{r registry}
list_learners()
```

```{r registry-tunable}
list_tunable_learners()
```

```{r registry-metrics}
list_metrics()
```

```{r registry-interpret}
list_interpretability_methods()
```

## Example data

This README uses `funcml::arthritis` as the main running example.

Here, `status` is the outcome for a binary classification task.

```{r demo-data}
demo_dat <- funcml::arthritis
demo_dat$status <- as.factor(demo_dat$status)
levels(demo_dat$status)
```

## Fit a classification model

`fit()` trains a model and returns a `funcml_fit` object.

```{r fit-model}
xgb_spec <- list(
  nrounds = 30,
  max_depth = 3,
  eta = 0.1,
  subsample = 1,
  colsample_bytree = 1
)

fit_obj <- fit(
  status ~ age + gender + bmi + diabetes + smoke + covered_health,
  data = demo_dat,
  model = "xgboost",
  spec = xgb_spec,
  seed = 42
)

fit_obj
```

## Generate predictions

The same fitted object can produce class predictions or class probabilities.

```{r predict-class}
predict(fit_obj, demo_dat[1:6, ])
```

```{r predict-prob}
pred_prob <- predict(
  fit_obj,
  demo_dat[1:6, ],
  type = "prob"
)

pred_prob
```

## Evaluate predictive performance

`evaluate()` applies the same learner under a resampling plan and returns fold-level results with summary statistics.

```{r evaluate-model}
eval_obj <- evaluate(
  data = demo_dat,
  formula = status ~ age + gender + bmi + diabetes + smoke + covered_health,
  model = "xgboost",
  spec = xgb_spec,
  resampling = cv(v = 4, seed = 42)
)

eval_obj
```

```{r evaluate-plot}
plot(eval_obj)
```

`funcml` also supports grouped cross-validation, time-based resampling, and holdout validation through `group_cv()`, `time_cv()`, and `holdout()`.

## Tune hyperparameters

`tune()` searches candidate hyperparameter settings using the same evaluation framework.

```{r tune-model}
tune_grid <- expand.grid(
  max_depth = c(2, 3),
  eta = c(0.05, 0.1),
  nrounds = c(20, 30)
)

tune_obj <- tune(
  data = demo_dat,
  formula = status ~ age + gender + bmi + diabetes + smoke + covered_health,
  model = "xgboost",
  grid = tune_grid,
  resampling = cv(v = 3, seed = 42),
  metric = "logloss",
  subsample = 1,
  colsample_bytree = 1,
  seed = 42
)

tune_obj
```

```{r tune-plot}
plot(tune_obj)
```

## Compare learners

`compare_learners()` benchmarks multiple learners under a common resampling design.

```{r compare-models}
compare_obj <- compare_learners(
  data = demo_dat,
  formula = status ~ age + gender + bmi + diabetes + smoke + covered_health,
  models = c("glm", "rpart", "xgboost"),
  metrics = c("accuracy", "logloss"),
  resampling = cv(v = 4, seed = 42),
  specs = list(xgboost = xgb_spec)
)

compare_obj
```

```{r compare-plot}
plot(compare_obj)
```

## Interpret fitted models

`interpret()` operates directly on fitted `funcml_fit` objects.

```{r interpret-permute}
permute_obj <- interpret(
  fit = fit_obj,
  data = demo_dat,
  method = "permute",
  nsim = 20,
  seed = 42
)

summary(permute_obj)
```

```{r interpret-permute-plot}
plot(permute_obj)
```

A second example shows accumulated local effects for one feature from the same fitted model.

```{r interpret-ale}
ale_obj <- interpret(
  fit = fit_obj,
  data = demo_dat,
  method = "ale",
  features = c("age"),
  type = "prob"
)

plot(ale_obj)
```

Other supported methods include PDP, ICE, SHAP, local explanations, surrogate models, interaction diagnostics, and calibration plots.

## Inspect calibration

For classification, the same interface also supports calibration diagnostics.

```{r calibration}
calibration_obj <- interpret(
  fit = fit_obj,
  data = demo_dat,
  method = "calibration",
  type = "prob",
  bins = 10,
  strategy = "quantile"
)

plot(calibration_obj)
```

## Estimate causal effects

`estimate()` extends the same framework to plug-in g-computation estimands such as the ATE.

The example below treats `smoke` as the treatment variable and `status` as the outcome, adjusting for the remaining covariates.

```{r estimate-causal}
est_obj <- estimate(
  data = demo_dat,
  formula = status ~ smoke + diabetes + age + gender + bmi + covered_health,
  model = "glm",
  estimand = "ATE",
  treatment = "smoke",
  interval = "normal",
  seed = 42
)

est_obj
```

```{r estimate-plot}
plot(est_obj)
```

The same interface also supports `ATT`, `CATE`, and `IATE`.

## Ensembles as first-class learners

Ensembles live in the same learner registry as base models.

```{r ensemble}
stack_fit <- fit(
  status ~ age + gender + bmi + diabetes + smoke + covered_health,
  data = demo_dat,
  model = "superlearner", # or "stacking"
  spec = list(
    learners = c("glm", "rpart", "xgboost", "nnet"),
    learner_specs = list(xgboost = xgb_spec),
    meta_model = "glmnet"
  ),
  seed = 42
)

predict(stack_fit, demo_dat[1:5, ], type = "prob")
```

## Summary

`funcml` provides a compact interface for tabular machine learning in R.

Use it to:

* train models
* generate predictions
* validate performance
* tune hyperparameters
* compare learners
* interpret fitted models
* estimate causal effects

The package is designed to keep the main analysis workflow explicit. 

## Contributing

Contributions are welcome.

For development setup, coding standards, and pull request guidelines, see `CONTRIBUTING.md`.

## Citation

If you use `funcml` in your work, cite the repository using GitHub's `Cite this repository` panel or the metadata in `CITATION.cff`.

APA:
```
El Badisy, I. (2026). funcml (Version 0.7.1) [Computer software]. https://github.com/ielbadisy/funcml
```

BibTeX:
```
@software{El_Badisy_funcml_2026, author = {El Badisy, Imad}, 
license = {GPL-3.0-only},
month = apr,
title = {{funcml}},
url = {https://github.com/ielbadisy/funcml},
version = {0.7.1},
year = {2026}
}
```
