---
output:
  github_document:
    fig_width: 8
    fig_height: 5
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  dpi = 120,
  warning = FALSE,
  message = FALSE
)
pkgload::load_all(".", export_all = FALSE, helpers = FALSE, quiet = TRUE)
ggplot2::theme_set(theme_funcml())
set.seed(42)
```

# funcml

`funcml` is a formula-first machine learning toolkit for people who want one
compact interface for fitting models, validating them, tuning them, comparing
learners, interpreting predictions, and estimating causal effects.

The package is intentionally opinionated: preprocessing happens before `fit()`,
the modeling input stays explicit, and the workflow remains compact instead of
expanding into a large orchestration framework.

## Why `funcml`?

- One consistent surface: `fit()`, `predict()`, `evaluate()`, `tune()`,
  `compare_learners()`, `interpret()`, and `estimate()`.
- Plot-ready outputs: evaluation, tuning, comparison, explanation,
  calibration, and treatment-effect objects all have native plot methods.
- Broad learner coverage: tree models, penalized GLMs, kernel methods,
  gradient boosting, Bayesian trees, and native ensemble learners.
- Audited learner layer: unsupported modes are blocked explicitly, and the
  repository includes the learner audit artifacts under
  [`work/audit/`](work/audit/).

## Installation

```r
install.packages("remotes")
remotes::install_github("ielbadisy/funcml")
```

`funcml` imports the supported learner backends directly, so a standard install
brings the learner registry into the package without extra registration steps.

## What the package covers

| Workflow | Main functions | Typical output |
| --- | --- | --- |
| Fit + predict | `fit()`, `predict()` | compact `funcml_fit` object with formula-aware prediction |
| Resampled validation | `cv()`, `holdout()`, `group_cv()`, `time_cv()`, `evaluate()` | fold-level metrics, uncertainty summaries, performance plots |
| Model selection | `tune()`, `compare_learners()` | search traces, tuned summaries, side-by-side learner comparison |
| Interpretation | `interpret()` | permutation importance, PDP, ICE, ALE, SHAP, surrogate, interaction, calibration |
| Causal estimation | `estimate()` | plug-in g-computation for `ATE`, `ATT`, `CATE`, and `IATE` |

## Quick start

```{r demo-data}
demo_reg <- transform(
  mtcars,
  car = rownames(mtcars)
)

demo_cls <- transform(
  mtcars,
  heavy = factor(ifelse(wt > median(wt), "heavy", "light"),
    levels = c("light", "heavy")
  )
)

demo_causal <- transform(
  demo_reg,
  trt = as.integer(wt > median(wt))
)

xgb_spec <- list(
  nrounds = 30,
  max_depth = 3,
  eta = 0.1,
  subsample = 1,
  colsample_bytree = 1
)
```

```{r quick-start}
fit_obj <- fit(
  mpg ~ wt + hp + qsec + drat,
  data = demo_reg,
  model = "xgboost",
  spec = xgb_spec,
  seed = 42
)

round(predict(fit_obj, demo_reg[1:5, , drop = FALSE]), 2)
```

The same fitted object can immediately feed evaluation, tuning,
interpretability, and causal workflows.

## Workflow gallery

### 1. Cross-validated model performance

```{r evaluate}
eval_obj <- evaluate(
  demo_reg,
  mpg ~ wt + hp + qsec + drat,
  model = "xgboost",
  spec = xgb_spec,
  resampling = cv(v = 4, seed = 42)
)

summary(eval_obj)
```

```{r eval-plot, fig.cap="Cross-validated regression performance from the same formula-first workflow."}
plot(eval_obj)
```

### 2. Tune a learner, then compare it against alternatives

```{r tune}
tune_grid <- expand.grid(
  max_depth = c(2, 3),
  eta = c(0.05, 0.1),
  nrounds = c(20, 30)
)

tune_obj <- tune(
  demo_reg,
  mpg ~ wt + hp + qsec + drat,
  model = "xgboost",
  grid = tune_grid,
  resampling = cv(v = 3, seed = 42),
  metric = "rmse",
  subsample = 1,
  colsample_bytree = 1,
  seed = 42
)

tune_obj$best
```

```{r tune-plot, fig.cap="Tuning trace for the xgboost workflow."}
plot(tune_obj)
```

```{r compare}
compare_obj <- compare_learners(
  demo_reg,
  mpg ~ wt + hp + qsec,
  models = c("glm", "rpart", "xgboost"),
  metrics = c("rmse", "mae"),
  resampling = cv(v = 4, seed = 42),
  specs = list(xgboost = xgb_spec)
)

summary(compare_obj)
```

```{r compare-plot, fig.cap="Direct learner comparison with uncertainty intervals."}
plot(compare_obj)
```

### 3. Build explanations from the same fitted model

```{r ale}
ale_obj <- interpret(
  fit_obj,
  demo_reg,
  method = "ale",
  features = c("wt", "hp")
)
```

```{r ale-plot, fig.cap="Accumulated local effects for two important predictors."}
plot(ale_obj)
```

```{r shap}
shap_obj <- interpret(
  fit_obj,
  demo_reg,
  method = "shap",
  newdata = demo_reg[1, , drop = FALSE],
  nsim = 30,
  nsamples = 20,
  seed = 42
)
```

```{r shap-plot, fig.cap="Local SHAP-style decomposition for one prediction."}
plot(shap_obj, kind = "waterfall")
```

The interpretation layer also supports permutation importance, PDP, ICE,
local surrogate explanations, global surrogates, interaction strength,
ceteris paribus profiles, and calibration diagnostics.

### 4. Inspect classification probabilities directly

```{r classification}
cls_fit <- fit(
  heavy ~ mpg + hp + qsec + drat,
  data = demo_cls,
  model = "glm",
  seed = 42
)

round(
  predict(cls_fit, demo_cls[1:6, , drop = FALSE], type = "prob"),
  3
)
```

```{r calibration}
calibration_obj <- interpret(
  cls_fit,
  demo_cls,
  method = "calibration",
  type = "prob",
  bins = 6
)
```

```{r calibration-plot, fig.cap="Probability calibration for a binary classification workflow."}
plot(calibration_obj)
```

### 5. Estimate causal effects with the same learner interface

```{r estimate}
est_obj <- estimate(
  demo_causal,
  mpg ~ trt + hp + qsec + drat,
  model = "rpart",
  estimand = "ATE",
  spec = list(cp = 0.01, minsplit = 5),
  interval = "bootstrap",
  n_boot = 20,
  seed = 42
)

summary(est_obj)
```

```{r estimate-plot, fig.cap="Estimated distribution of unit-level treatment effects from plug-in g-computation."}
plot(est_obj)
```

## Ensembles are first-class learners

`stacking` and `superlearner` live in the same registry as the base learners,
so you fit them through the same `fit()` interface.

```{r ensemble}
stack_fit <- fit(
  mpg ~ wt + hp + qsec + drat,
  data = demo_reg,
  model = "stacking",
  spec = list(
    learners = c("glm", "rpart", "xgboost"),
    learner_specs = list(xgboost = xgb_spec),
    meta_model = "glmnet"
  ),
  seed = 42
)

round(predict(stack_fit, demo_reg[1:5, , drop = FALSE]), 2)
```

## Learner coverage

The current registry covers 26 learner ids. Broadly:

| Support | Learners |
| --- | --- |
| Regression + classification | `glm`, `rpart`, `glmnet`, `ranger`, `nnet`, `e1071_svm`, `randomForest`, `gbm`, `kknn`, `ctree`, `cforest`, `lightgbm`, `catboost`, `xgboost`, `stacking`, `superlearner` |
| Regression + binary classification | `earth`, `gam`, `bart` |
| Classification only | `C50`, `naivebayes`, `fda`, `lda`, `qda` |
| Binary classification only | `adaboost` |
| Regression only | `pls` |

The learner adapter layer was audited package-wide. Unsupported combinations
are now rejected early instead of failing later in evaluation or
interpretation code.

## Local development note

- `README.md` is generated from `README.Rmd`.
- The audit-backed learner status artifacts live under [`work/audit/`](work/audit/).
- The full `tests/testthat` suite currently passes against the source tree.
