---
output:
  github_document:
    fig_width: 8
    fig_height: 5
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 120,
  warning = FALSE,
  message = FALSE
)
pkgload::load_all(".", export_all = FALSE, helpers = FALSE, quiet = TRUE)
ggplot2::theme_set(theme_funcml())
set.seed(42)
```

# funcml

`funcml` is a machine learning framework for R with one compact interface for
fitting models, validating them, tuning them, comparing learners,
interpreting predictions, and estimating causal effects.

The package is intentionally opinionated: preprocessing happens before `fit()`,
the modeling input stays explicit, and the framework stays compact instead of
expanding into a large orchestration framework.

## Why `funcml`?

- One consistent surface: `fit()`, `predict()`, `evaluate()`, `tune()`,
  `compare_learners()`, `interpret()`, and `estimate()`.
- Plot-ready outputs: evaluation, tuning, comparison, explanation,
  calibration, and treatment-effect objects all have native plot methods.
- Broad learner coverage: classical statistical models, trees, kernel methods,
  boosting systems, Bayesian trees, and native ensemble learners.
- Audited learner layer: unsupported modes are blocked explicitly.

## Installation

```r
install.packages("remotes")
remotes::install_github("ielbadisy/funcml")
```

Installing `funcml` makes the learner registry available without extra
registration steps.

## API overview

| Task | Main functions | Returned object | Typical use |
| --- | --- | --- | --- |
| Model fitting | `fit()`, `predict()` | `funcml_fit` | Train one learner and generate predictions |
| Resampled validation | `cv()`, `holdout()`, `group_cv()`, `time_cv()`, `evaluate()` | `funcml_eval` | Estimate out-of-sample performance with uncertainty |
| Model selection | `tune()`, `compare_learners()` | `funcml_tune`, `funcml_compare` | Search hyperparameters and compare learners under common resampling |
| Interpretation | `interpret()` | method-specific interpretation classes | Explain fitted models with global and local diagnostics |
| Causal estimation | `estimate()` | `funcml_estimand` | Estimate plug-in g-computation estimands |

## Demo data used below

The README uses one regression problem, one binary classification problem, and
one synthetic causal example so the same API can be shown across the package
surface.

```{r demo-data}
demo_reg <- transform(
  mtcars,
  car = rownames(mtcars)
)

demo_cls <- local({
  x1 <- rnorm(500)
  x2 <- rnorm(500)
  x3 <- runif(500, -1, 1)
  eta <- -0.4 + 1.0 * x1 - 0.8 * x2 + 0.7 * x3
  data.frame(
    outcome = factor(
      ifelse(runif(500) < stats::plogis(eta), "yes", "no"),
      levels = c("no", "yes")
    ),
    x1 = x1,
    x2 = x2,
    x3 = x3
  )
})

demo_cls_test <- local({
  x1 <- rnorm(250)
  x2 <- rnorm(250)
  x3 <- runif(250, -1, 1)
  eta <- -0.4 + 1.0 * x1 - 0.8 * x2 + 0.7 * x3
  data.frame(
    outcome = factor(
      ifelse(runif(250) < stats::plogis(eta), "yes", "no"),
      levels = c("no", "yes")
    ),
    x1 = x1,
    x2 = x2,
    x3 = x3
  )
})

demo_causal <- local({
  x1 <- rnorm(600)
  x2 <- rnorm(600)
  x3 <- runif(600, -1, 1)
  p_trt <- stats::plogis(-0.2 + 0.7 * x1 - 0.5 * x2 + 0.4 * x3)
  trt <- rbinom(600, 1, p_trt)
  true_effect <- 1.2 + 0.6 * x3
  outcome <- 3 + true_effect * trt + 0.8 * x1 - 0.7 * x2 + 0.5 * x3 +
    rnorm(600, sd = 0.4)
  data.frame(
    outcome = outcome,
    trt = trt,
    x1 = x1,
    x2 = x2,
    x3 = x3,
    true_effect = true_effect
  )
})

xgb_spec <- list(
  nrounds = 30,
  max_depth = 3,
  eta = 0.1,
  subsample = 1,
  colsample_bytree = 1
)
```

## 1. Start by inspecting the learner catalog

`list_learners()` gives a session-aware inventory of the registry, including
task support, probability support, multiclass support, engine packages, and
whether each engine is currently available.

```{r learner-catalog}
catalog <- list_learners()

catalog_summary <- data.frame(
  quantity = c(
    "Total learners",
    "Regression support",
    "Classification support",
    "Probability support",
    "Multiclass support",
    "Engines currently available"
  ),
  value = c(
    nrow(catalog),
    sum(catalog$supports_regression),
    sum(catalog$supports_classification),
    sum(catalog$supports_prob),
    sum(catalog$supports_multiclass),
    sum(catalog$available)
  )
)

catalog_summary
```

```{r learner-catalog-preview}
catalog[, c(
  "learner",
  "supports_regression",
  "supports_classification",
  "supports_prob",
  "supports_multiclass",
  "engine_package",
  "available"
)]
```

## 2. Fit one model and inspect the fitted object

The entry point is `fit()`. It returns a compact `funcml_fit` object that
stores the learner id, formula, encoded feature layout, and prediction
machinery.

```{r fit-model}
fit_obj <- fit(
  mpg ~ wt + hp + qsec + drat,
  data = demo_reg,
  model = "xgboost",
  spec = xgb_spec,
  seed = 42
)

fit_obj
```

```{r fit-summary}
summary(fit_obj)
```

```{r fit-predict}
data.frame(
  car = demo_reg$car[1:6],
  observed = demo_reg$mpg[1:6],
  predicted = round(predict(fit_obj, demo_reg[1:6, , drop = FALSE]), 2)
)
```

## 3. Validate performance with resampling

`evaluate()` applies the same learner interface under a resampling plan and
returns fold-level results plus uncertainty summaries.

```{r evaluate}
eval_obj <- evaluate(
  demo_reg,
  mpg ~ wt + hp + qsec + drat,
  model = "xgboost",
  spec = xgb_spec,
  resampling = cv(v = 4, seed = 42)
)

summary(eval_obj)
```

```{r eval-plot}
plot(eval_obj)
```

The same resampling interface can also handle grouped CV, rolling time splits,
or plain holdout validation through `group_cv()`, `time_cv()`, and `holdout()`.

## 4. Tune hyperparameters, then compare learners

`tune()` searches a grid or random sample of hyperparameters using the same
evaluation machinery. `compare_learners()` then places multiple learners under
the same resampling design for an apples-to-apples comparison.

```{r tune}
tune_grid <- expand.grid(
  max_depth = c(2, 3),
  eta = c(0.05, 0.1),
  nrounds = c(20, 30)
)

tune_obj <- tune(
  demo_reg,
  mpg ~ wt + hp + qsec + drat,
  model = "xgboost",
  grid = tune_grid,
  resampling = cv(v = 3, seed = 42),
  metric = "rmse",
  subsample = 1,
  colsample_bytree = 1,
  seed = 42
)

tune_obj$best[, c(names(tune_grid), "mean", "conf_low", "conf_high")]
```

```{r tune-table}
tune_obj$results[
  order(tune_obj$results$mean),
  c(names(tune_grid), "mean", "conf_low", "conf_high")
]
```

```{r tune-plot}
plot(tune_obj)
```

```{r compare}
compare_obj <- compare_learners(
  demo_reg,
  mpg ~ wt + hp + qsec,
  models = c("glm", "rpart", "xgboost"),
  metrics = c("rmse", "mae"),
  resampling = cv(v = 4, seed = 42),
  specs = list(xgboost = xgb_spec)
)

summary(compare_obj)
```

```{r compare-plot}
plot(compare_obj)
```

## 5. Interpret a fitted model with the same object

The interpretation layer operates on fitted `funcml_fit` objects, so you do not
need a second modeling interface for explanation tasks.

```{r permute}
permute_obj <- interpret(
  fit_obj,
  demo_reg,
  method = "permute",
  nsim = 20,
  seed = 42
)

summary(permute_obj)
```

```{r permute-plot}
plot(permute_obj)
```

```{r ale}
ale_obj <- interpret(
  fit_obj,
  demo_reg,
  method = "ale",
  features = c("wt", "hp")
)
```

```{r ale-plot}
plot(ale_obj)
```

```{r shap}
shap_obj <- interpret(
  fit_obj,
  demo_reg,
  method = "shap",
  newdata = demo_reg[1, , drop = FALSE],
  nsim = 30,
  nsamples = 20,
  seed = 42
)
```

```{r shap-plot}
plot(shap_obj, kind = "waterfall")
```

Other supported methods include PDP, ICE, local surrogate explanations, global
surrogates, interaction strength, ceteris paribus profiles, and calibration
diagnostics.

## 6. Inspect class probabilities and calibration

For classification, the same `fit()` object can produce class predictions or
class-probability matrices with aligned columns.

```{r classification}
cls_fit <- fit(
  outcome ~ x1 + x2 + x3,
  data = demo_cls,
  model = "glm",
  seed = 42
)

cls_prob <- round(
  predict(cls_fit, demo_cls_test[1:6, , drop = FALSE], type = "prob"),
  3
)

data.frame(
  prob_no = cls_prob[, "no"],
  prob_yes = cls_prob[, "yes"],
  row.names = NULL
)
```

```{r calibration}
calibration_obj <- interpret(
  cls_fit,
  demo_cls_test,
  method = "calibration",
  type = "prob",
  bins = 8,
  strategy = "quantile"
)
```

```{r calibration-plot}
plot(calibration_obj)
```

## 7. Estimate causal effects with the same learner interface

`estimate()` extends the same framework into plug-in g-computation for common
binary-treatment estimands.

This synthetic causal example has measured confounding and a known treatment
effect centered near `1.2`, so the ATE output has a meaningful target.

```{r estimate}
est_obj <- estimate(
  demo_causal,
  outcome ~ trt + x1 + x2 + x3 + trt:x3,
  model = "glm",
  estimand = "ATE",
  treatment_level = 1,
  control_level = 0,
  interval = "normal",
  seed = 42
)

summary(est_obj)
```

```{r estimate-plot}
plot(est_obj)
```

The same API also supports `ATT`, `CATE`, and `IATE`.

## 8. Use ensembles as first-class learners

`stacking` and `superlearner` live in the same learner registry as the base
models, so ensembles are fit through the same API rather than a separate
pipeline.

```{r ensemble}
stack_fit <- fit(
  mpg ~ wt + hp + qsec + drat,
  data = demo_reg,
  model = "stacking",
  spec = list(
    learners = c("glm", "rpart", "xgboost"),
    learner_specs = list(xgboost = xgb_spec),
    meta_model = "glmnet"
  ),
  seed = 42
)

round(predict(stack_fit, demo_reg[1:5, , drop = FALSE]), 2)
```

## 9. Learner coverage at a glance

The current registry covers `r nrow(catalog)` learner ids. Broadly:

| Support | Learners |
| --- | --- |
| Regression + classification | `glm`, `rpart`, `glmnet`, `ranger`, `nnet`, `e1071_svm`, `randomForest`, `gbm`, `kknn`, `ctree`, `cforest`, `lightgbm`, `xgboost`, `stacking`, `superlearner` |
| Regression + binary classification | `earth`, `gam`, `bart` |
| Classification only | `C50`, `naivebayes`, `fda`, `lda`, `qda` |
| Binary classification only | `adaboost` |
| Regression only | `pls` |

## 10. Summary

`funcml` is designed so the same fitted object and the same learner registry can
support:

- one-model training with `fit()`
- prediction with `predict()`
- resampled validation with `evaluate()`
- hyperparameter search with `tune()`
- learner benchmarking with `compare_learners()`
- model explanation with `interpret()`
- plug-in g-computation with `estimate()`

That is the central idea of the package: one explicit API surface for tabular
machine learning in R, rather than a stack of disconnected modeling wrappers.
