---
output:
  github_document:
    fig_width: 8
    fig_height: 5
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  dpi = 120,
  warning = FALSE,
  message = FALSE
)
pkgload::load_all(".", export_all = FALSE, helpers = FALSE, quiet = TRUE)
ggplot2::theme_set(theme_funcml())
set.seed(42)
```

# funcml

`funcml` is a machine learning framework for R with one explicit interface for
fitting models, validating them, tuning them, comparing learners,
interpreting predictions, and estimating causal effects.

The package is intentionally opinionated: preprocessing happens before
modeling, inputs stay explicit, and the API stays compact instead of expanding
into a large orchestration framework.

## Why `funcml`?

- One surface for the full modeling loop: `fit()`, `predict()`, `evaluate()`,
  `tune()`, `compare_learners()`, `interpret()`, and `estimate()`.
- Session-aware learner catalog via `list_learners()`, including capability and
  availability metadata.
- Plot-ready outputs across validation, tuning, comparison, explanation,
  calibration, and treatment-effect workflows.
- Native support for stacked and super learner ensembles through the same
  interface as base learners.

## Installation

```r
install.packages("remotes")
remotes::install_github("ielbadisy/funcml")
```

After installation, inspect the learner catalog with `list_learners()`. This
shows which learner ids are exposed through the compact `funcml` API and which
backends are available in the current R session.

## API Overview

| Task | Main functions | Returned object | Typical use |
| --- | --- | --- | --- |
| Learner discovery | `list_learners()` | `data.frame` | Inspect learner ids, capabilities, and engine availability |
| Model fitting | `fit()`, `predict()` | `funcml_fit` | Train one learner and generate predictions |
| Resampled validation | `cv()`, `holdout()`, `group_cv()`, `time_cv()`, `evaluate()` | `funcml_eval` | Estimate out-of-sample performance with uncertainty |
| Model selection | `tune()`, `compare_learners()` | `funcml_tune`, `funcml_compare` | Search hyperparameters and compare learners under common resampling |
| Interpretation | `interpret()` | method-specific interpretation classes | Explain fitted models with global and local diagnostics |
| Causal estimation | `estimate()` | `funcml_estimand` | Estimate plug-in g-computation estimands |

## Demo data used below

The README uses one regression problem, one binary classification problem, and
one synthetic causal example so the same API can be shown across the package
surface.

```{r demo-data}
demo_reg <- transform(
  mtcars,
  car = rownames(mtcars)
)

demo_cls <- local({
  x1 <- rnorm(500)
  x2 <- rnorm(500)
  x3 <- runif(500, -1, 1)
  eta <- -0.4 + 1.0 * x1 - 0.8 * x2 + 0.7 * x3
  data.frame(
    outcome = factor(
      ifelse(runif(500) < stats::plogis(eta), "yes", "no"),
      levels = c("no", "yes")
    ),
    x1 = x1,
    x2 = x2,
    x3 = x3
  )
})

demo_cls_test <- local({
  x1 <- rnorm(250)
  x2 <- rnorm(250)
  x3 <- runif(250, -1, 1)
  eta <- -0.4 + 1.0 * x1 - 0.8 * x2 + 0.7 * x3
  data.frame(
    outcome = factor(
      ifelse(runif(250) < stats::plogis(eta), "yes", "no"),
      levels = c("no", "yes")
    ),
    x1 = x1,
    x2 = x2,
    x3 = x3
  )
})

demo_causal <- local({
  x1 <- rnorm(600)
  x2 <- rnorm(600)
  x3 <- runif(600, -1, 1)
  p_trt <- stats::plogis(-0.2 + 0.7 * x1 - 0.5 * x2 + 0.4 * x3)
  trt <- rbinom(600, 1, p_trt)
  true_effect <- 1.2 + 0.6 * x3
  outcome <- 3 + true_effect * trt + 0.8 * x1 - 0.7 * x2 + 0.5 * x3 +
    rnorm(600, sd = 0.4)
  data.frame(
    outcome = outcome,
    trt = trt,
    x1 = x1,
    x2 = x2,
    x3 = x3,
    true_effect = true_effect
  )
})

xgb_spec <- list(
  nrounds = 30,
  max_depth = 3,
  eta = 0.1,
  subsample = 1,
  colsample_bytree = 1
)
```

## 1. Inspect the learner catalog

`list_learners()` now follows a compact registry style by default.

```{r learner-catalog}
detailed_catalog <- list_learners(
  columns = c(
    "learner",
    "has_tune",
    "supports_regression",
    "supports_classification",
    "supports_prob",
    "supports_multiclass",
    "supports_importance",
    "engine_package",
    "available"
  )
)

catalog <- list_learners()

catalog
```

```{r learner-catalog-summary}
data.frame(
  learners = nrow(detailed_catalog),
  tune = sum(detailed_catalog$has_tune),
  regression = sum(detailed_catalog$supports_regression),
  classification = sum(detailed_catalog$supports_classification),
  prob = sum(detailed_catalog$supports_prob),
  multiclass = sum(detailed_catalog$supports_multiclass),
  importance = sum(detailed_catalog$supports_importance),
  available = sum(detailed_catalog$available)
)
```

```{r learner-catalog-preview}
list_tunable_learners()
```

```{r learner-id-preview}
head(list_learners(
  classification = TRUE,
  prob = TRUE,
  columns = c("learner", "has_tune", "supports_prob", "supports_multiclass", "engine_package")
), 6)
```

```{r interpretability-catalog}
list_interpretability_methods()
```

```{r metrics-catalog}
list_metrics()
```

## 2. Fit one model and inspect the fitted object

`fit()` is the entry point for training a single learner. It returns a compact
`funcml_fit` object that stores the learner id, formula, encoded feature
layout, and prediction machinery.

```{r fit-model}
fit_obj <- fit(
  mpg ~ wt + hp + qsec + drat,
  data = demo_reg,
  model = "xgboost",
  spec = xgb_spec,
  seed = 42
)

fit_obj
```

```{r fit-predict}
head(data.frame(
  car = demo_reg$car[1:6],
  observed = demo_reg$mpg[1:6],
  pred = round(predict(fit_obj, demo_reg[1:6, , drop = FALSE]), 2)
), 6)
```

## 3. Validate performance with resampling

`evaluate()` applies the same learner interface under a resampling plan and
returns fold-level results plus uncertainty summaries.

```{r evaluate}
eval_obj <- evaluate(
  data = demo_reg,
  formula = mpg ~ wt + hp + qsec + drat,
  model = "xgboost",
  spec = xgb_spec,
  resampling = cv(v = 4, seed = 42)
)

eval_tbl <- eval_obj$summary[, c("metric", "mean", "conf_low", "conf_high")]
eval_tbl[, c("mean", "conf_low", "conf_high")] <- round(
  eval_tbl[, c("mean", "conf_low", "conf_high")],
  3
)
eval_tbl
```

```{r eval-plot}
plot(eval_obj)
```

The same resampling interface also handles grouped CV, rolling time splits,
and plain holdout validation through `group_cv()`, `time_cv()`, and
`holdout()`.

## 4. Tune hyperparameters and compare learners

`tune()` searches a grid or random sample of hyperparameters using the same
evaluation machinery. `compare_learners()` then puts multiple learners under
the same resampling design for an apples-to-apples comparison.

```{r tune}
tune_grid <- expand.grid(
  max_depth = c(2, 3),
  eta = c(0.05, 0.1),
  nrounds = c(20, 30)
)

tune_obj <- tune(
  data = demo_reg,
  formula = mpg ~ wt + hp + qsec + drat,
  model = "xgboost",
  grid = tune_grid,
  resampling = cv(v = 3, seed = 42),
  metric = "rmse",
  subsample = 1,
  colsample_bytree = 1,
  seed = 42
)

round(tune_obj$best[, c(names(tune_grid), "mean", "conf_low", "conf_high")], 3)
```

```{r tune-table}
tune_tbl <- head(
  tune_obj$results[
    order(tune_obj$results$mean),
    c(names(tune_grid), "mean", "conf_low", "conf_high")
  ],
  4
)
tune_tbl[, c("mean", "conf_low", "conf_high")] <- round(
  tune_tbl[, c("mean", "conf_low", "conf_high")],
  3
)
tune_tbl
```

```{r tune-plot}
plot(tune_obj)
```

```{r compare}
compare_obj <- compare_learners(
  data = demo_reg,
  formula = mpg ~ wt + hp + qsec,
  models = c("glm", "rpart", "xgboost"),
  metrics = c("rmse", "mae"),
  resampling = cv(v = 4, seed = 42),
  specs = list(xgboost = xgb_spec)
)

compare_tbl <- compare_obj$results[, c("model", "metric", "mean", "conf_low", "conf_high", "rank")]
compare_tbl[, c("mean", "conf_low", "conf_high")] <- round(
  compare_tbl[, c("mean", "conf_low", "conf_high")],
  3
)
compare_tbl
```

```{r compare-plot}
plot(compare_obj)
```

## 5. Interpret a fitted model

The interpretation layer operates directly on fitted `funcml_fit` objects, so
you do not need a second modeling interface for explanation tasks.

```{r permute}
permute_obj <- interpret(
  fit = fit_obj,
  data = demo_reg,
  method = "permute",
  nsim = 20,
  seed = 42
)

summary(permute_obj)
```

```{r permute-plot}
plot(permute_obj)
```

```{r ale}
ale_obj <- interpret(
  fit = fit_obj,
  data = demo_reg,
  method = "ale",
  features = c("wt", "hp")
)
```

```{r ale-plot}
plot(ale_obj)
```

```{r shap}
shap_obj <- interpret(
  fit = fit_obj,
  data = demo_reg,
  method = "shap",
  newdata = demo_reg[1, , drop = FALSE],
  nsim = 30,
  nsamples = 20,
  seed = 42
)
```

```{r shap-plot}
plot(shap_obj, kind = "waterfall")
```

Other supported methods include PDP, ICE, local surrogate explanations, global
surrogates, interaction strength, ceteris paribus profiles, and calibration
diagnostics.

## 6. Inspect class probabilities and calibration

For classification, the same `fit()` object can produce class predictions or
class-probability matrices with aligned columns.

```{r classification}
cls_fit <- fit(
  outcome ~ x1 + x2 + x3,
  data = demo_cls,
  model = "glm",
  seed = 42
)

cls_prob <- round(
  predict(cls_fit, demo_cls_test[1:6, , drop = FALSE], type = "prob"),
  3
)

data.frame(
  prob_no = cls_prob[, "no"],
  prob_yes = cls_prob[, "yes"],
  row.names = NULL
)
```

```{r calibration}
calibration_obj <- interpret(
  fit = cls_fit,
  data = demo_cls_test,
  method = "calibration",
  type = "prob",
  bins = 8,
  strategy = "quantile"
)
```

```{r calibration-plot}
plot(calibration_obj)
```

## 7. Estimate causal effects through the same interface

`estimate()` extends the same framework into plug-in g-computation for common
binary-treatment estimands.

This synthetic causal example has measured confounding and a known treatment
effect centered near `1.2`, so the ATE output has a meaningful target.

```{r estimate}
est_obj <- estimate(
  data = demo_causal,
  formula = outcome ~ trt + x1 + x2 + x3 + trt:x3,
  model = "glm",
  estimand = "ATE",
  treatment_level = 1,
  control_level = 0,
  interval = "normal",
  seed = 42
)

data.frame(
  estimand = est_obj$estimand,
  treatment_level = est_obj$treatment_level,
  control_level = est_obj$control_level,
  estimate = round(est_obj$estimate, 3),
  conf_low = round(est_obj$conf_int[1], 3),
  conf_high = round(est_obj$conf_int[2], 3)
)
```

```{r estimate-plot}
plot(est_obj)
```

The same API also supports `ATT`, `CATE`, and `IATE`.

## 8. Use ensembles as first-class learners

`stacking` and `superlearner` live in the same catalog as the base learners, so
ensembles are fit through the same API rather than a separate pipeline.

```{r ensemble}
stack_fit <- fit(
  mpg ~ wt + hp + qsec + drat,
  data = demo_reg,
  model = "stacking",
  spec = list(
    learners = c("glm", "rpart", "xgboost"),
    learner_specs = list(xgboost = xgb_spec),
    meta_model = "glmnet"
  ),
  seed = 42
)

round(predict(stack_fit, demo_reg[1:5, , drop = FALSE]), 2)
```

## 9. Capability summary from the live catalog

Rather than hardcoding learner names, you can summarize the current registry
directly from `list_learners()`.

```{r catalog-summary}
catalog_summary <- data.frame(
  capability = c(
    "Supports regression",
    "Supports classification",
    "Supports probabilities",
    "Supports multiclass",
    "Supports feature importance",
    "Currently available"
  ),
  learners = c(
    paste(detailed_catalog$learner[detailed_catalog$supports_regression], collapse = ", "),
    paste(detailed_catalog$learner[detailed_catalog$supports_classification], collapse = ", "),
    paste(detailed_catalog$learner[detailed_catalog$supports_prob], collapse = ", "),
    paste(detailed_catalog$learner[detailed_catalog$supports_multiclass], collapse = ", "),
    paste(detailed_catalog$learner[detailed_catalog$supports_importance], collapse = ", "),
    paste(detailed_catalog$learner[detailed_catalog$available], collapse = ", ")
  ),
  row.names = NULL
)

catalog_summary
```

## 10. Summary

`funcml` is designed so the same fitted object model and the same learner
catalog can support:

- learner discovery with `list_learners()`
- one-model training with `fit()`
- prediction with `predict()`
- resampled validation with `evaluate()`
- hyperparameter search with `tune()`
- learner benchmarking with `compare_learners()`
- model explanation with `interpret()`
- plug-in g-computation with `estimate()`

That is the central idea of the package: one explicit API surface for tabular
machine learning in R, rather than a stack of disconnected modeling wrappers.
