---
output:
  github_document:
    fig_width: 8
    fig_height: 5
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.align = "center",
  dpi = 120
)
pkgload::load_all(".", export_all = FALSE, helpers = FALSE, quiet = TRUE)
set.seed(42)
```

# funcml

`funcml` is a formula-first functional machine learning package for building
end-to-end modeling workflows with a compact S3 interface.

The package is intentionally opinionated. It is not meant to compete feature
for feature with more established and more complete frameworks such as
`tidymodels`, `mlr3`, or `caret`. The goal is narrower: provide a compact,
explicit, formula-first pipeline for users who want to fit, compare, tune,
interpret, and estimate machine learning models without adopting a large
workflow system.

The package is centered on one workflow:

- `fit()`
- `predict()`
- `learners()`
- `compare_learners()`
- `cv()`
- `evaluate()`
- `tune()`
- `estimate()`
- `interpret()`

Interpretability is included as one part of the package, not the main purpose.

## Design philosophy

`funcml` is built around a deliberate tradeoff: less framework machinery in
exchange for a more compact, explicit workflow.

- Preprocessing is intentionally outside the package. Data modifications should
  happen before `fit()` so the modeling input is explicit and the training step
  does not hide transformation state.
- The package favors functional composition and clear inputs over implicit
  workflow objects that mutate or learn preprocessing steps inside training.
- Cross-validation is still the default scientific baseline, but the package
  now also supports holdout splits, grouped CV, and time-aware rolling
  evaluation through the same resampling interface.
- This is a tradeoff, not a universal claim. Users who need richer
  preprocessing orchestration, broader resampling strategies, or larger ML
  ecosystems should use more complete frameworks.

In short, `funcml` is for users who want the most compact pipeline that still
keeps model fitting, evaluation, tuning, interpretability, and causal
estimation in one coherent interface.

## Package context

`funcml` is organized around a single learner registry and a shared S3 modeling
surface.

- `fit()` encodes a formula/data pair, dispatches to a learner adapter, and
  returns a `funcml_fit` object with a stored prediction closure.
- `predict()` reuses the stored design-matrix metadata so new data must match
  the fitted feature space exactly.
- `evaluate()` and `tune()` build on the same `fit()` interface.
- `compare_learners()` compares several learners on a shared resampling scheme,
  with optional tuning inside the comparison step.
- `estimate()` performs plug-in g-computation for causal estimands.
- `interpret()` provides native model-agnostic explanations and plot methods.

Current learner ids exposed through `learners()` include:

| Task support | Learners |
| --- | --- |
| Regression + classification | `glm`, `rpart`, `glmnet`, `ranger`, `nnet`, `e1071_svm`, `randomForest`, `gbm`, `kknn`, `ctree`, `cforest`, `lightgbm`, `catboost`, `xgboost`, `stacking`, `superlearner` |
| Regression + binary classification | `earth`, `gam`, `bart` |
| Classification only | `C50`, `naivebayes`, `fda`, `lda`, `qda` |
| Binary classification only | `adaboost` |
| Regression only | `pls` |

## Current local package status

In this workspace, the source tree is ahead of the older installed package that
`library(funcml)` may pick up from a user library. The source version includes
the expanded learner registry for `gam`, `bart`, `ctree`, `cforest`,
`naivebayes`, `fda`, `adaboost`, and `pls`. This README is rendered from the
current source tree, not the stale installed copy.

## Full API demonstration with `xgboost`

The example below uses one learner, `xgboost`, across the full exported API.

```{r demo-data}
demo_data <- transform(
  mtcars,
  trt = as.integer(wt > median(wt)),
  car = rownames(mtcars)
)

xgb_spec <- list(
  nrounds = 40,
  max_depth = 3,
  eta = 0.1,
  subsample = 1,
  colsample_bytree = 1
)

learners()[grepl("xgboost", learners())]
```

### Fit, predict, and basic metrics

```{r fit-predict}
fit_obj <- fit(
  mpg ~ wt + hp + qsec + drat,
  data = demo_data,
  model = "xgboost",
  spec = xgb_spec,
  seed = 42
)

fit_obj
summary(fit_obj)
```

```{r predictions}
pred <- predict(fit_obj, demo_data[1:5, , drop = FALSE])
round(pred, 3)
```

```{r metrics}
fitted_pred <- predict(fit_obj, demo_data)

data.frame(
  metric = c("rmse", "mae", "rsq"),
  value = c(
    rmse(demo_data$mpg, fitted_pred),
    mae(demo_data$mpg, fitted_pred),
    rsq(demo_data$mpg, fitted_pred)
  )
)
```

### Cross-validation and evaluation

```{r evaluate}
resampling <- cv(v = 4, repeats = 2, seed = 42)

eval_obj <- evaluate(
  demo_data,
  mpg ~ wt + hp + qsec + drat,
  model = "xgboost",
  spec = xgb_spec,
  resampling = resampling
)

eval_obj
summary(eval_obj)
```

The summary now includes fold-level standard errors and confidence intervals.

```{r evaluate-other-resampling}
holdout_obj <- evaluate(
  demo_data,
  mpg ~ wt + hp + qsec + drat,
  model = "glm",
  resampling = holdout(prop = 0.8, seed = 42)
)

time_obj <- evaluate(
  transform(demo_data, order_id = seq_len(nrow(demo_data))),
  mpg ~ wt + hp + qsec + drat,
  model = "glm",
  resampling = time_cv(initial = 16, assess = 4, time = "order_id")
)

holdout_obj$summary
time_obj$summary
```

```{r eval-plot, fig.cap="Cross-validated performance for the xgboost regression workflow."}
plot(eval_obj)
```

### Hyperparameter tuning

```{r tune}
tune_grid <- expand.grid(
  max_depth = c(2, 3),
  eta = c(0.05, 0.1),
  nrounds = c(20, 40)
)

tune_obj <- tune(
  demo_data,
  mpg ~ wt + hp + qsec + drat,
  model = "xgboost",
  grid = tune_grid,
  resampling = cv(v = 4, seed = 42),
  metric = "rmse",
  subsample = 1,
  colsample_bytree = 1
)

tune_obj
summary(tune_obj)
```

Budgeted random search is also supported for larger grids.

```{r tune-random}
random_tune_obj <- tune(
  demo_data,
  mpg ~ wt + hp + qsec + drat,
  model = "xgboost",
  grid = tune_grid,
  search = "random",
  n_evals = 3,
  resampling = cv(v = 4, seed = 42),
  metric = "rmse",
  subsample = 1,
  colsample_bytree = 1,
  seed = 42
)

random_tune_obj$best
```

```{r tune-plot, fig.cap="Grid-search trace for xgboost."}
plot(tune_obj)
```

### Causal effect estimation

`estimate()` uses the same learner interface. Here the treatment is the binary
indicator `trt`.

```{r estimate}
est_obj <- estimate(
  demo_data,
  mpg ~ trt + hp + qsec + drat,
  model = "xgboost",
  estimand = "ATE",
  spec = xgb_spec,
  interval = "bootstrap",
  n_boot = 30,
  seed = 42
)

est_obj
summary(est_obj)
```

```{r estimate-plot, fig.cap="Distribution of estimated unit-level treatment effects from plug-in g-computation."}
plot(est_obj)
```

### Native interpretability workflow

#### Permutation importance

```{r permute}
permute_obj <- interpret(
  fit_obj,
  demo_data,
  method = "permute",
  metric = "rmse",
  nsim = 10,
  seed = 42
)

permute_obj
summary(permute_obj)
```

```{r permute-plot, fig.cap="Permutation importance for the xgboost model."}
plot(permute_obj)
```

#### Partial dependence

```{r pdp}
pdp_obj <- interpret(
  fit_obj,
  demo_data,
  method = "pdp",
  features = c("wt", "hp")
)

pdp_obj
summary(pdp_obj)
```

```{r pdp-plot, fig.cap="Partial dependence profiles for `wt` and `hp`."}
plot(pdp_obj)
```

#### ICE curves

```{r ice}
ice_obj <- interpret(
  fit_obj,
  demo_data,
  method = "ice",
  features = "wt",
  nsamples = 20
)

ice_obj
summary(ice_obj)
```

```{r ice-plot, fig.cap="ICE curves for `wt`."}
plot(ice_obj)
```

#### ALE

```{r ale}
ale_obj <- interpret(
  fit_obj,
  demo_data,
  method = "ale",
  features = c("wt", "hp")
)

ale_obj
summary(ale_obj)
```

```{r ale-plot, fig.cap="Accumulated local effects for `wt` and `hp`."}
plot(ale_obj)
```

#### Local surrogate explanation

```{r local-model}
local_obj <- interpret(
  fit_obj,
  demo_data,
  method = "local_model",
  newdata = demo_data[1, , drop = FALSE],
  nsamples = 20,
  k = 3
)

local_obj
summary(local_obj)
```

```{r local-plot, fig.cap="Top local surrogate contributions for the first row."}
plot(local_obj)
```

#### SHAP approximation

```{r shap}
shap_obj <- interpret(
  fit_obj,
  demo_data,
  method = "shap",
  newdata = demo_data[1, , drop = FALSE],
  nsim = 30,
  nsamples = 20,
  seed = 42
)

shap_obj
summary(shap_obj)
```

```{r shap-plot, fig.cap="SHAP waterfall plot for the first row."}
plot(shap_obj, kind = "waterfall")
```

The SHAP interface also supports additional plot kinds:

- `plot(shap_obj, kind = "summary")` for a SHAP summary / beeswarm view
  when explanations are computed for multiple rows
- `plot(shap_obj, kind = "importance")` or `kind = "bar"` for mean absolute
  SHAP importance
- `plot(shap_obj, kind = "auto")` to choose waterfall for one row and summary
  for multiple rows

```{r shap-other-types, eval=FALSE}
shap_multi <- interpret(
  fit_obj,
  demo_data,
  method = "shap",
  newdata = demo_data[1:10, , drop = FALSE],
  nsim = 30,
  nsamples = 20,
  seed = 42
)

plot(shap_multi, kind = "summary")
plot(shap_multi, kind = "importance")
plot(shap_multi, kind = "auto")
```

#### Global surrogate

```{r surrogate}
surrogate_obj <- interpret(
  fit_obj,
  demo_data,
  method = "surrogate",
  maxdepth = 2
)

surrogate_obj
summary(surrogate_obj)
```

```{r surrogate-plot, fig.cap="Global surrogate model view."}
plot(surrogate_obj)
```

#### Interaction strength

```{r interaction}
interaction_obj <- interpret(
  fit_obj,
  demo_data,
  method = "interaction",
  nsamples = 20,
  grid_size = 8
)

interaction_obj
summary(interaction_obj)
```

```{r interaction-plot, fig.cap="Estimated interaction strength by feature."}
plot(interaction_obj)
```

## Comparing multiple learners

`compare_learners()` compares several models directly, or tunes each one first
and then compares the tuned best configurations on one or more metrics. The
result summaries also include uncertainty intervals around the resampled means.

```{r compare-learners}
compare_obj <- compare_learners(
  demo_data,
  mpg ~ wt + hp + qsec,
  models = c("glm", "rpart", "xgboost"),
  metrics = c("rmse", "mae"),
  resampling = cv(v = 4, seed = 42),
  specs = list(
    xgboost = list(nrounds = 20, max_depth = 3, eta = 0.1, subsample = 1, colsample_bytree = 1)
  )
)

compare_obj
summary(compare_obj)
```

```{r compare-learners-plot, fig.cap="Side-by-side learner comparison across multiple metrics."}
plot(compare_obj)
```

```{r compare-learners-tuned}
compare_tuned_obj <- compare_learners(
  demo_data,
  mpg ~ wt + hp + qsec,
  models = c("rpart", "xgboost"),
  tune = TRUE,
  metric = "rmse",
  metrics = c("rmse", "mae"),
  grids = list(
    rpart = expand.grid(cp = c(0.001, 0.01), minsplit = c(5, 10)),
    xgboost = expand.grid(max_depth = c(2, 3), eta = c(0.05, 0.1), nrounds = c(10, 20))
  ),
  resampling = cv(v = 4, seed = 42),
  subsample = 1,
  colsample_bytree = 1
)

compare_tuned_obj
summary(compare_tuned_obj)
```

## Notes on local development

- `testthat::test_local(".")` passes against the source tree.
- A direct `library(funcml)` call may still resolve to an older installed copy
  until the package is reinstalled from this source checkout.
