---
title: "funcml"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{funcml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(funcml)
```

`funcml` provides a formula-first workflow for fitting and evaluating models.
The package also includes native interpretability helpers for permutation
importance, partial dependence, ICE, ALE, local surrogate explanations, SHAP,
interaction strength, greedy breakdown profiles, and global surrogate models.
It also exposes native ensemble learners through `model = "stacking"` and
`model = "superlearner"`.

```{r}
fit_obj <- fit(mpg ~ wt + hp, data = mtcars, model = "glm")

permute_obj <- interpret(fit_obj, mtcars, method = "permute", nsim = 5)
pdp_obj <- interpret(fit_obj, mtcars, method = "pdp", features = "wt")
ale_obj <- interpret(fit_obj, mtcars, method = "ale", features = "wt")
local_obj <- interpret(fit_obj, mtcars, method = "local_model", newdata = mtcars[1, , drop = FALSE], k = 2)
shap_obj <- interpret(fit_obj, mtcars, method = "shap", newdata = mtcars[1, , drop = FALSE], nsim = 20)
interaction_obj <- interpret(fit_obj, mtcars, method = "interaction")
breakdown_obj <- interpret(fit_obj, mtcars, method = "breakdown", newdata = mtcars[1, , drop = FALSE])
surrogate_obj <- interpret(fit_obj, mtcars, method = "surrogate")
```

```{r}
plot(permute_obj)
plot(pdp_obj)
plot(ale_obj)
plot(local_obj)
plot(shap_obj, kind = "waterfall")
plot(interaction_obj)
plot(breakdown_obj)
plot(surrogate_obj)
```
