---
title: "funcml"
output: rmarkdown::pdf_document
vignette: >
  %\VignetteIndexEntry{funcml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(funcml)
```

`funcml` provides a machine learning framework for R. The package also includes native
interpretability helpers for permutation
importance, partial dependence, ICE, ALE, local surrogate explanations, SHAP,
interaction strength, greedy breakdown profiles, and global surrogate models.
It also exposes native ensemble learners through `model = "stacking"` and
`model = "superlearner"`.

The package is intentionally opinionated. It is not designed to compete
feature for feature with broader frameworks such as `tidymodels`, `mlr3`, or
`caret`. Instead, `funcml` focuses on a compact and explicit framework for
users who want to fit, compare, tune, interpret, and estimate models through
one coherent interface, with formula syntax as the main model-specification
surface.

This design comes with a deliberate tradeoff:

- preprocessing is expected to happen outside `funcml`, so the data passed to
  `fit()` is the exact data being modeled
- the package favors explicit inputs and direct model specification over hidden
  preprocessing state inside training pipelines
- cross-validation is still the default, but the package also supports holdout
  splits, grouped CV, and time-aware rolling evaluation through the same
  resampling interface

Users who need broader preprocessing orchestration or specialized resampling
designs should use a more complete framework.

Learner coverage currently includes:

- regression and classification: `glm`, `rpart`, `glmnet`, `ranger`, `nnet`, `e1071_svm`, `randomForest`, `gbm`, `kknn`, `ctree`, `cforest`, `lightgbm`, `xgboost`, `stacking`, `superlearner`
- regression and binary classification: `earth`, `gam`, `bart`
- classification only: `C50`, `naivebayes`, `fda`, `lda`, `qda`
- binary classification only: `adaboost`
- regression only: `pls`

```{r}
fit_obj <- fit(mpg ~ wt + hp, data = mtcars, model = "glm")

permute_obj <- interpret(fit_obj, mtcars, method = "permute", nsim = 5)
pdp_obj <- interpret(fit_obj, mtcars, method = "pdp", features = "wt")
ale_obj <- interpret(fit_obj, mtcars, method = "ale", features = "wt")
local_obj <- interpret(fit_obj, mtcars, method = "local_model", newdata = mtcars[1, , drop = FALSE], k = 2)
shap_obj <- interpret(fit_obj, mtcars, method = "shap", newdata = mtcars[1, , drop = FALSE], nsim = 20)
profile_obj <- interpret(fit_obj, mtcars, method = "profile", newdata = mtcars[1, , drop = FALSE])
surrogate_obj <- interpret(fit_obj, mtcars, method = "surrogate")
```

```{r}
eval_obj <- evaluate(mpg ~ wt + hp, data = mtcars, model = "glm", resampling = holdout(prop = 0.8, seed = 1))
eval_obj$summary
```

```{r}
tune_grid <- expand.grid(intercept = c(TRUE, FALSE))
tune_obj <- tune(mpg ~ wt + hp, data = mtcars, model = "glm", grid = tune_grid, search = "random", n_evals = 1, resampling = cv(v = 3, seed = 1), seed = 1)
tune_obj$best
```

Nested CV is available through `outer_resampling`. The inner `resampling`
argument still selects the best configuration, and the outer resampling loop
provides an unbiased estimate of tuned model-selection performance.

```{r}
nested_tune_obj <- tune(
  mpg ~ wt + hp,
  data = mtcars,
  model = "glm",
  grid = tune_grid,
  resampling = cv(v = 3, seed = 1),
  outer_resampling = cv(v = 4, seed = 2),
  metric = "rmse",
  seed = 1
)
nested_tune_obj$nested$summary
```

```{r, eval = FALSE}
interaction_obj <- interpret(fit_obj, mtcars, method = "interaction")
plot(interaction_obj)
```

```{r}
learners()[1:10]
```

```{r}
plot(permute_obj)
plot(pdp_obj)
plot(ale_obj)
plot(local_obj)
plot(shap_obj, kind = "waterfall")
plot(profile_obj)
plot(surrogate_obj)
```
