---
title: "funcml"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{funcml}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(funcml)
```

`funcml` provides a formula-first workflow for fitting and evaluating models.
The package also includes native interpretability helpers for permutation
importance, partial dependence, ICE, ALE, local surrogate explanations, SHAP,
interaction strength, greedy breakdown profiles, and global surrogate models.
It also exposes native ensemble learners through `model = "stacking"` and
`model = "superlearner"`.

The package is intentionally opinionated. It is not designed to compete
feature for feature with broader frameworks such as `tidymodels`, `mlr3`, or
`caret`. Instead, `funcml` focuses on a compact, explicit, formula-first
workflow for users who want to fit, compare, tune, interpret, and estimate
models through one coherent interface.

This design comes with a deliberate tradeoff:

- preprocessing is expected to happen outside `funcml`, so the data passed to
  `fit()` is the exact data being modeled
- the package favors functional composition and explicit inputs over hidden
  preprocessing state inside training workflows
- cross-validation is still the default, but the package also supports holdout
  splits, grouped CV, and time-aware rolling evaluation through the same
  resampling interface

Users who need broader preprocessing orchestration or specialized resampling
designs should use a more complete framework.

Learner coverage currently includes:

- regression and classification: `glm`, `rpart`, `glmnet`, `ranger`, `nnet`, `e1071_svm`, `randomForest`, `gbm`, `kknn`, `ctree`, `cforest`, `lightgbm`, `catboost`, `xgboost`, `stacking`, `superlearner`
- regression and binary classification: `earth`, `gam`, `bart`
- classification only: `C50`, `naivebayes`, `fda`, `lda`, `qda`
- binary classification only: `adaboost`
- regression only: `pls`

```{r}
fit_obj <- fit(mpg ~ wt + hp, data = mtcars, model = "glm")

permute_obj <- interpret(fit_obj, mtcars, method = "permute", nsim = 5)
pdp_obj <- interpret(fit_obj, mtcars, method = "pdp", features = "wt")
ale_obj <- interpret(fit_obj, mtcars, method = "ale", features = "wt")
local_obj <- interpret(fit_obj, mtcars, method = "local_model", newdata = mtcars[1, , drop = FALSE], k = 2)
shap_obj <- interpret(fit_obj, mtcars, method = "shap", newdata = mtcars[1, , drop = FALSE], nsim = 20)
profile_obj <- interpret(fit_obj, mtcars, method = "profile", newdata = mtcars[1, , drop = FALSE])
surrogate_obj <- interpret(fit_obj, mtcars, method = "surrogate")
```

```{r}
eval_obj <- evaluate(mpg ~ wt + hp, data = mtcars, model = "glm", resampling = holdout(prop = 0.8, seed = 1))
eval_obj$summary
```

```{r}
tune_grid <- expand.grid(intercept = c(TRUE, FALSE))
tune_obj <- tune(mpg ~ wt + hp, data = mtcars, model = "glm", grid = tune_grid, search = "random", n_evals = 1, resampling = cv(v = 3, seed = 1), seed = 1)
tune_obj$best
```

```{r, eval = FALSE}
interaction_obj <- interpret(fit_obj, mtcars, method = "interaction")
plot(interaction_obj)
```

```{r}
learners()[1:10]
```

```{r}
plot(permute_obj)
plot(pdp_obj)
plot(ale_obj)
plot(local_obj)
plot(shap_obj, kind = "waterfall")
plot(profile_obj)
plot(surrogate_obj)
```
